---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Code for reading in the dataset and/or processing the data
=============================================================

```{r}
#Download the zipfile
#first makesure that the directory data exists
if (!file.exists("data")){
     dir.create("data")
}
###This code is for Checking whether data file exists. If not, it will unzip the file
#setwd("./data") # move to the directory of the data
#https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

if (!file.exists("./data/NEI_data.zip")) { 
     fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
     download.file(fileUrl, destfile = "./data/repdata_data_activity.zip",method="curl")
     unzip("./data/repdata_data_activity.zip",exdir="./data") #unzip the file
}
```

Then read in the activity file
```{r}
library(data.table)
ActivityMonitoringData <- fread("./data/activity.csv")
#format date as date
ActivityMonitoringData$date <- as.Date(ActivityMonitoringData$date,"%Y-%m-%d")

```



## What is mean total number of steps taken per day?

Total number of steps taken each day
========================================================

```{r}
#Histogram
#aggregate per day
StepsperDay <- aggregate(ActivityMonitoringData$steps, by=list(ActivityMonitoringData$date), sum)
names(StepsperDay) <- c("Date","Steps")
str(StepsperDay)

#Histogram number of steps per day
library(ggplot2)
histogram <- ggplot(StepsperDay,aes(Date,Steps))
histogram <- histogram + geom_bar(stat="identity")

print(histogram)
```

Mean and median number of steps taken each day
=====
Because on a number of days the steps are 'NA', they have to be removed before the Mean and Median are calculated.
```{r}
mean(StepsperDay$Steps,na.rm=TRUE)
median(StepsperDay$Steps,na.rm=TRUE)

```

## What is the average daily activity pattern?

Time series plot of the average number of steps taken
===

First average the number of steps per 5-minute interval; remove the NA's when calculating the mean.

```{r}
#aggregate per interval
StepsperInterval <- aggregate(ActivityMonitoringData$steps, 
                    by=list(ActivityMonitoringData$interval), mean, na.rm=TRUE)
names(StepsperInterval) <- c("Interval","Average_Steps")
head(StepsperInterval)
```

Then make a time series plot of it

```{r}
#Timeseries average steps per interval
library(ggplot2)
ts_plot <- ggplot(StepsperInterval,aes(Interval,Average_Steps))
ts_plot <- ts_plot + geom_line()
print(ts_plot)
```



The 5-minute interval that, on average, contains the maximum number of steps
===

To find the interval with the maximum number of average steps, we use the function which.max()
```{r}
StepsperInterval[which.max(StepsperInterval$Average_Steps),]

```

## Imputing missing values

Code to describe and show a strategy for imputing missing data
===
First find out how many NAs there are, and how many in the column Steps
```{r}
sum(is.na(ActivityMonitoringData))
sum(is.na(ActivityMonitoringData$steps))
```
So there are only missing values in the column steps.


There are 2 strategies dor imputing missing data:

1. Remove the missing values. This is done in the previous steps.

2. Replace the missing values by 0; which in this case could make sense.

In this case: first replace the missing values by zero.

```{r}
ActivityMonitoringDataZero <- ActivityMonitoringData
ActivityMonitoringDataZero[is.na(ActivityMonitoringDataZero)] <- 0

```


Then you re-do the graphs before with the zeroed dataset.

Histogram of the total number of steps taken each day after missing values are imputed
===

```{r}
#Histogram
#aggregate per day
StepsperDayZero <- aggregate(ActivityMonitoringDataZero$steps, by=list(ActivityMonitoringDataZero$date), sum)
names(StepsperDayZero) <- c("Date","Steps")

#Histogram number of steps per day
library(ggplot2)
histogramZero <- ggplot(StepsperDayZero,aes(Date,Steps))
histogramZero <- histogramZero + geom_bar(stat="identity")

print(histogramZero)
```

Then also calculate the Mean and Median, that shows the bias that was caused by leaving out the NAs:
```{r}
mean(StepsperDayZero$Steps)
median(StepsperDayZero$Steps)

```


## Are there differences in activity patterns between weekdays and weekends?
Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
===

First make a column with the day of the week, and a factor column to indicate whether it is a weekday.

```{r}
ActivityMonitoringDataZero$Weekday <- weekdays(ActivityMonitoringDataZero$date)
ActivityMonitoringDataZero$WeekdayFactor <- ifelse(
     ActivityMonitoringDataZero$Weekday %in% c("zaterdag","zondag","saturday","sunday"),"Weekend", "Weekday")

```

Please note: this is using Dutch and English language settings; you may want to change that depending on your own language settings

Calculate the average steps per interval, for weekdays and weekends

```{r}
#aggregate per interval and weekdayindicator
StepsperIntervalWeekday <- aggregate(ActivityMonitoringDataZero$steps, 
     by=list(ActivityMonitoringDataZero$WeekdayFactor,ActivityMonitoringDataZero$interval), 
     mean)
names(StepsperIntervalWeekday) <- c("Weekdayfactor","Interval","Average_Steps")
head(StepsperIntervalWeekday)

```

Then create a panel plot 


```{r}
#Timeseries average steps per interval
library(ggplot2)
ts_plot <- ggplot(StepsperIntervalWeekday,aes(Interval,Average_Steps))
ts_plot <- ts_plot + geom_line()
ts_plot <- ts_plot + facet_grid(Weekdayfactor ~ ., scales = "free") # separate the weekday types
print(ts_plot)
```
